Pareto-Optimal Multistate Protein Design using Tree Graphical Models

### 2012-12-06 docs/tex/draft.tex
Completed psuedo-code of quickhull
Added clarification on energy convex combination Added references. Described quickhull. Started writing CMRF convex.

\subsection{Multistate protein design}
The problem of protein structure prediction is finding the lowest energy structure given a particular sequence. Protein design is known as the inverse protein structure prediction problem and involves finding the lowest energy sequence for a given structure. Multistate protein design generalizes protein design by asking to find a sequence that is stable for multiple structures. This is often posed as a  multi-objective optimization problem \ref{}.

@@

\subsubsection*{Quick hull procedure}
The Quick hull procedure is used to find all the points on the pareto optimal frontier. The quick hull is traditionally used to find the convex hull of a set of points. In our case we are only interested in the lower left convex frontier. The algorithm in pseudo code is described in Algorithm~\ref{algo:quickhull1}.

\begin{algorithm}
\caption{Quick Hull} \label{algo:quickhull1}
\begin{algorithmic}
\For{each position $i$}
	\For{each rotamer $r$ at $i$}
		\For{each candidate rotamer $t$ at $i$}
			\State $X \gets E_i(r_i^r) - E_i(r_i^t)$
			\ForAll{positions $j \not= i$} 
				\ForAll{rotamers $u$ at position $j$}
					\State $Y = \min_u\left[E_{ij}(r_i^r,r_j^u) - E_{ij}(r_i^t,r_j^u) \right]$
				\EndFor
				\State $X \gets X + Y$
			\EndFor
			\If{$X>0$}
				\State eliminate $r_i^r$
				\State break
			\EndIf
		\EndFor
	\EndFor
\EndFor
\end{algorithmic}
\end{algorithm}

@@

\subsection{Convex combinations of CMRFs}

\begin{figure}[h]
% selecting unit length
\centering
% used for centering Figure
\setlength{\unitlength}{0.3cm}
\begin{picture}(40,25)(0,0)
\multiput(5,15)(8,0){2}{\circle{4}}
\multiput(5,7)(8,0){2}{\circle{4}}
\multiput(5,7)(8,0){2}{\circle{3.7}}
\multiput(5,23)(8,0){2}{\circle{4}}
\multiput(5,23)(8,0){2}{\circle{3.7}}
\thicklines
\multiput(7,15)(8,0){2}{\line(1,0){4}}
\multiput(20,15)(1,0){4}{\circle*{0.3}}
\multiput(5,13)(8,0){2}{\line(0,-1){4}}
\multiput(5,17)(8,0){2}{\line(0,1){4}}
\put(24,15){\line(1,0){4}}
\put(30,13){\line(0,-1){4}}
\put(30,17){\line(0,1){4}}
\thinlines
\put(30,15){\circle{4}}
\put(30,7){\circle{4}}
\put(30,7){\circle{3.7}}
\put(30,23){\circle{4}}
\put(30,23){\circle{3.7}}
\put(3.5,14.6){$\phi(X_1)$}
\put(11.5,14.6){$\phi(X_2)$}
\put(28.5,14.6){$\phi(X_n)$}
\put(3.5,6.6){$\phi(O^A_1)$}
\put(11.5,6.6){$\phi(O^A_2)$}
\put(28.5,6.6){$\phi(O^A_n)$}
\put(3.5,22.6){$\phi(O^B_1)$}
\put(11.5,22.6){$\phi(O^B_2)$}
\put(28.5,22.6){$\phi(O^B_n)$}
\put(7,16.5){\footnotesize{$\psi(X_1{,}X_2)$}}
\put(15,16.5){\footnotesize{$\psi(X_2{,}X_3)$}}
\put(22.5,16.5){\footnotesize{$\psi(X_{n{-}1}{,}X_n)$}}
\put(5.2,10.6){\footnotesize{$\psi(X_1{,}O^A_1)$}}
\put(13.2,10.6){\footnotesize{$\psi(X_2{,}O^A_2)$}}
\put(30.2,10.6){\footnotesize{$\psi(X_n{,}O^A_n)$}}
\put(5.2,18.6){\footnotesize{$\psi(X_1{,}O^B_1)$}}
\put(13.2,18.6){\footnotesize{$\psi(X_2{,}O^B_2)$}}
\put(30.2,18.6){\footnotesize{$\psi(X_n{,}O^B_n)$}}
\end{picture}
\label{fig:CMRF2}
\caption{A Tree structured Markov Random Field that is used for modelling multi-state protein design.}
\end{figure}

@@

>>>>

\subsection{Multistate protein design}
The problem of protein structure prediction is finding the lowest energy structure given a particular sequence. Protein design is known as the inverse protein structure prediction problem and involves finding the lowest energy sequence for a given structure. Multistate protein design generalizes protein design by asking to find a sequence that is stable for multiple structures. This is often posed as a  multi-objective optimization problem~\cite{Zheng2009}.

@@

@@

\subsubsection*{Quick hull procedure}
The Quick hull procedure is used to find all the points on the pareto optimal frontier. The quick hull is traditionally used to find the convex hull of a set of points. In our case we are only interested in the lower left convex frontier similar to ~\cite{Zheng2009}.  The algorithm in pseudo code is described in Algorithm~\ref{algo:quickhull1}. The basic idea behind the algorithm is to find a pair of points on the pareto optimal frontier. Find the point farthest from the line joining the two points and then to repeat the procedure with this newly found point and the previous two points. When there are no newer points to find, all points on the pareto optimal frontier have been discovered.

\begin{algorithm}
\caption{Quick Hull} \label{algo:quickhull1}
\begin{algorithmic}
\State Initialize queue $Q \gets \emptyset$
\State Find $X_A \gets \arg \min_X E_A(X) $
\State Find $X_B \gets \arg \min_X E_B(X) $
\State Enqueue $(X_A,X_B)$
\State $C_H \gets \{X_A,X_B\}$
\Repeat
	\State Dequeue from Q a pair $(X_1,X_2)$ and assert($E_A(X_1) < E_A(X_2)$)
	\State $m \gets \frac{E_B(X_1)-E_B(X_2)}{E_A(X_1)-E_A(X_2)}$
	\State $\theta_A \gets \frac{-m}{1-m}$
	\State $\theta_B \gets \frac{1}{1-m}$
	\State Find $X_{AB} \gets \arg \min_X \theta_A E_A(X) + \theta_B E_B(X) $
	\If{$X_{AB} \neq X_A$ and $X_{AB} \neq X_B$ }
		\State $C_H \gets C_H \cup \{X_{AB}\}$
		\State Enqueue $(X_A,X_{AB})$ and $(X_{AB},X_B)$
	\EndIf
\Until{Q is empty}
\State \Return $C_H$
\end{algorithmic}
\end{algorithm}

\subsection{Convex combinations of CMRFs}
For multi-state protein design we can combine the energy functions as follows.
\[
E_{AB}(X) = \theta_A E_A(X) + \theta_B E_B(X) \quad \text{where} \quad \theta_A,\theta_B \geq 0 \text{ and } \theta_A + \theta_B =1
\]
Here $E_A(X)$ is any energy function mapping sequence to energy values for state A. The lower the energy the better the fit for sequence X to state A. $E_A(X)$ can also be written as a CMRF.
\[
E_A(X) = -\log  \left[{\prod_{i=1}^{n}{\phi(X_i)\phi(\mathbf{O^A_i})\psi(\mathbf{O^A_i},X_i})\prod_{i=1}^{n-1}{\psi(X_i,X_{i+1})}}\right]
\]
Note that $-\log$ is to formulate the problem as a minimization problem so that it agrees with the energy concept. However the optimal sequence remain the same. 
\[
\begin{split}
X_A &= \arg \min_X -\log  \left[{\prod_{i=1}^{n}{\phi(X_i)\phi(\mathbf{O^A_i})\psi(\mathbf{O^A_i},X_i})\prod_{i=1}^{n-1}{\psi(X_i,X_{i+1})}}\right]\\
&= \arg \max_X \left[{\prod_{i=1}^{n}{\phi(X_i)\phi(\mathbf{O^A_i})\psi(\mathbf{O^A_i},X_i})\prod_{i=1}^{n-1}{\psi(X_i,X_{i+1})}}\right]
\end{split}
\]
Taking the $-\log$ also offers benefits as it lets us combine convex combinations of CMRF energy functions and gives rise to new graphical models called Tree MRFS (TMRFs). These TMRFs are conceptually very similar to CMRFs and possess the properties of efficient inference and decoding.

\begin{align*}
X_A &= \arg \min_X -\log  \left[{\prod_{i=1}^{n}{\phi(X_i)\phi(\mathbf{O^A_i})\psi(\mathbf{O^A_i},X_i})\prod_{i=1}^{n-1}{\psi(X_i,X_{i+1})}}\right]
\end{align*}

\begin{figure}[h!]
% selecting unit length
\centering
% used for centering Figure
\setlength{\unitlength}{0.3cm}
\begin{picture}(40,25)(0,0)
\multiput(5,15)(8,0){2}{\circle{4}}
\multiput(5,7)(8,0){2}{\circle{4}}
\multiput(5,7)(8,0){2}{\circle{3.7}}
\multiput(5,23)(8,0){2}{\circle{4}}
\multiput(5,23)(8,0){2}{\circle{3.7}}
\thicklines
\multiput(7,15)(8,0){2}{\line(1,0){4}}
\multiput(20,15)(1,0){4}{\circle*{0.3}}
\multiput(5,13)(8,0){2}{\line(0,-1){4}}
\multiput(5,17)(8,0){2}{\line(0,1){4}}
\put(24,15){\line(1,0){4}}
\put(30,13){\line(0,-1){4}}
\put(30,17){\line(0,1){4}}
\thinlines
\put(30,15){\circle{4}}
\put(30,7){\circle{4}}
\put(30,7){\circle{3.7}}
\put(30,23){\circle{4}}
\put(30,23){\circle{3.7}}
\put(3.5,14.6){$\phi(X_1)$}
\put(11.5,14.6){$\phi(X_2)$}
\put(28.5,14.6){$\phi(X_n)$}
\put(3.5,6.6){$\phi(O^A_1)$}
\put(11.5,6.6){$\phi(O^A_2)$}
\put(28.5,6.6){$\phi(O^A_n)$}
\put(3.5,22.6){$\phi(O^B_1)$}
\put(11.5,22.6){$\phi(O^B_2)$}
\put(28.5,22.6){$\phi(O^B_n)$}
\put(7,16.5){\footnotesize{$\psi(X_1{,}X_2)$}}
\put(15,16.5){\footnotesize{$\psi(X_2{,}X_3)$}}
\put(22.5,16.5){\footnotesize{$\psi(X_{n{-}1}{,}X_n)$}}
\put(5.2,10.6){\footnotesize{$\psi(X_1{,}O^A_1)$}}
\put(13.2,10.6){\footnotesize{$\psi(X_2{,}O^A_2)$}}
\put(30.2,10.6){\footnotesize{$\psi(X_n{,}O^A_n)$}}
\put(5.2,18.6){\footnotesize{$\psi(X_1{,}O^B_1)$}}
\put(13.2,18.6){\footnotesize{$\psi(X_2{,}O^B_2)$}}
\put(30.2,18.6){\footnotesize{$\psi(X_n{,}O^B_n)$}}
\end{picture}
\label{fig:CMRF2}
\caption{A Tree structured Markov Random Field that is used for modelling multi-state protein design.}
\end{figure}

@@

\subsubsection{Clarification on Energy Convex Combination}
From the discussion in the beginning, the reader may have been lead to believe that we are taking convex combinations of the likelihood of multiple states and then maximizing this likelihood. While in spirit this is true, however practically we are optimizing something else. Let $L(X|O^A)$  and $L(X|O^B)$ be the likelihoods of the sequences under state A and state B respectively. The reader may have been lead to believe that we are solving the following problem
\begin{equation}
\label{eqn:clarif1}
X^* = \arg \max_X \left( \theta_A L(X|O^A) + \theta_B L(X|O^B) \right)
\end{equation}
But in, reality we are solving : 
\begin{align}
X^* &= \arg \min_X \left( -\theta_A \log L(X|O^A) - \theta_B \log L(X|O^B) \right) \\
&= \arg \min_X \left( \log \frac{1}{L(X|O^A)^{\theta_A} L(X|O^B)^{\theta_B}} \right) \\
&= \arg \max_X \left( L(X|O^A)^{\theta_A} L(X|O^B)^{\theta_B} \right)
\label{eqn:clarif2} 
\end{align}

Clearly Eqn~\ref{eqn:clarif2} is not the same as Eqn~\ref{eqn:clarif1}. One is minimizing the arithmetic mean, while the other is minimizing the geometric mean. It remains to be seen what the semantic consequence of this mathematical detail is.
